Improving the modelling of tempo and phoneme duration

### 2017-05-22 Paper/Paper_v1.tex
Added 2 notebooks on neat analysis of vowels and consonants.

\documentclass[11pt,a4paper]{scrbook}
\usepackage{geometry}	
\usepackage[backend=bibtex,style=numeric]{biblatex}
\usepackage{ngerman}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[style=numeric]{biblatex}
\bibliography{../PaperBib}
\usepackage{colortbl}	
\usepackage{xcolor}
\usepackage{soul}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[outdir=../Graphen/]{epstopdf}
\usepackage{textcomp}
\usepackage{floatrow}
\renewcommand{\figurename}{Fig.}
\usepackage{caption}
\setcapindent{0pt}
\usepackage{subcaption}
\usepackage{scrhack}
\renewcommand{\familydefault}{\sfdefault}
\definecolor{uhhred}{cmyk}{0,100,100,0}
\usepackage{tabularx}

@@

\section{Method}
Our starting premise was that phoneme durations are being influenced by the speaking tempo. In addition to this, we assumed that this influence is not equal on all phonemes, and on all situations, and therefore also not linear. So we formulated our aim as to investigate the influence of speech rate on the duration of different phonemes and find a way of using this information to improve the performance of phoneme prediction models.

In order to achieve this aim, we first examined the phonemes with respect to their duration. We found out that the classical phonetics book grouping of phonemes is not respected in terms of duration. There are consonants like \texttt{/N/}, and \texttt{/m/} which may be longer than some long vowels, and long vowels which are shorter than some short vowels or even several consonants, like \texttt{/y:/}. If one would apply a clustering method on phonemes to group them on seven classes based on their duration interval coordinates using, say k-means method, the resulting groups would certainly differ from those listed in Chapter 4.

Therefore we decided to capture phoneme identity in our model as well, although it inflates the number of attributes and attribute values to compute. The relation between phoneme duration and factors other than speech rate was treated only marginally, as these have already been extensively examined in the literature on various corpora. For the German language, intuition aside, we have information on such factors provided by: Kohler \cite{Kohler1992}, Riedi \cite{Riedi1995}, Moebius \cite{Moebius1996}, and Brinckmann \& Trouvain \cite{Brinckmann_2003}, who applied some of the existent segment duration models to the German language.At this stage, we considered such factors only to the extent they might have helped identify groups of phonemes that may be best suited to evaluate the influence of speech rate. Such a factor is primary stress, which creates a group of very elastic vowels, which are therefore also sensitive to changes in speech tempo.

Then we looked for a definition of speech rate that can be used to model phoneme durations. This supposed the presence of a good correlation between speech rate and phoneme duration, at least for some phoneme groups. We explained our choice whith respect to this in Chapter 5. 
The fact, that speech rate doesn't show any correlation with some phonemes, like \texttt{/r/}, while it has a pretty good correlation with others, like \texttt{/a:/} suggests the need for classification in whichever model we choose.

@@

\begin{enumerate}
	\item The first one predicted the phoneme duration as the mean of its durations across the whole corpus. 
	\item The second one build a two level tree, which split the root on the 49 realized phonemes, and then each phoneme on two specific speech rate values. The leaves contained three predicted values for each phoneme. These values were calculated using the mean duration of the specific phoneme across the corpus modified or not by a proportion of its standard deviation, based on the speech rate value split. 
\end{enumerate}

@@

Next, we pruned the used feature space by using two feature selection algorithms: forward and backward feature selection. Their principle is easy to understand and they are available in Weka. Consequently, following features were selected:

@@

	\begin{itemize}
		\item phoneme id: all the 52 phoneme definitions used in Verbmobil
		\item phoneme type: vowel, or consonant
		\item articulation manner: vowel, plosive, fricative, nasal, lateral, approximant, other
	\end{itemize}

@@

	\begin{itemize}
		\item position of phoneme in syllable: initial, final
		\item syllable part containing current phoneme: onset, nucleus, code, single-phoneme-syllable
		\item lexical stress of the syllable (using the stress type of the nucleus vowel)
	\end{itemize}

@@

	\begin{itemize}
		\item previous phoneme type: vowel, consonant, or none
		\item following phoneme type: vowel, or consonant, or none
		\item articulation manner of following phoneme: vowel, plosive, fricative, nasal, lateral, approximant, other, none
		\item if following phoneme is voiced: yes, no, or none
		\item syllable part to which following phoneme belongs: onset, nucleus, code, single-phoneme-syllable
	\end{itemize}
\end{itemize}

Note that all features consider only actually realised units in speech. We also had to reduce the amount of data used for the model to 58\% of the total corpus, due to capacity limitations imposed by Weka on our system.
After pruning, the \textit{word} related features got completely eliminated, together with \textit{syllable-place-in-word} and \textit{articulation-of-realised-previous phoneme}.

To this collection of features we added \textit{speech rate} in the second phase and compare the results.

\section{Results}
We trained and evaluated both a M5P and a REPTree model. The results were very similar, so present here only the results obtained using the model tree M5P, as they were slightly better. As can be seen in Table \ref{tab:perfM} adding the speech rate feature caused a significant improvement on all our models.

@@

\hline
	 & Basic - SR & Basic + SR & M5P - SR & M5P + SR & CART\\
\hline
\hline
Correlation coef 		& .42  &  & .50 & .79 & .86 / .83\\
	\hline
MAE  &  31.02 & & 34.21 & 25.19 & 22.46 / 21.40\\
	\hline
RMSE  &  47.09 & & 62.91 & 44.70 & NA \\
	\hline
\end{tabular}
\caption{Overview of the performance results of 3 models with (+) and without (-) speech rate (SR). CART is the model created by Brinckmann \& Trouvain \cite{Brinckmann_2003}. The two values in the CART cells represent the performance obtained for 2 different speakers.} 
\label{tab:perfM}
\end{table}

However, the performance of our model tree is smaller than the performance of the CART built by Brinckmann \& Trouvain \cite{Brinckmann_2003}. This is explainable if we consider the nature of the used datasets. The CART tree used PhonDat, a rather homogenous dataset in terms of speech rate and number of speakers. Furthermore, the model was trained and tested separately on data produced by one speaker. The reported results show even for PhonDat a difference in performance between speakers. Another aspect that can have a great impact on the model result is the quality of the input data. While PhonDat was manually segmented and annotated, Verbmobil was processed automatically using the MAUS segmentation system. When evaluating outliers we found many examples of wrong segmentation, as we also discussed in Section 4.2.

On the other hand, our model tree shows robustness with respect to factors like inter-speaker variability and speech rate. The results obtained for a test set containing data from a speaker not included in the training data were similar to those obtained on known speakers on terms of correlation coefficient, and showed less error rates, as can be seen in table \ref{tab:perfM5P}. Moreover, if the model is to be evaluated on data produced by only one speaker, it produces good results even when using less than 10\% of the amount of data used to train on the multiple speaker corpus. The most frequent speaker in our database still only has 185 recorded turns, consisting of 9315 identifyiable uttered phonemes, which represent 0.02\% of the total phoneme instances in our corpus, so we could not test our model on a larger dataset of a single speaker. In comparison to this, the dataset used for CART contained 624 sentences, and a total of 4932 orthographically different words, which results in more than a double sized corpus.

@@

\end{tabular}
\caption{Overview of the performance of our M5P model tree on different training and test sets.} 
\label{tab:perfM5P}
\end{table}

\chapter{Conclusion and Outlook}
Our model calculates an approximation for the phoneme duration when its corresponding speech segment changes in terms of duration. It does this no matter what was the cause of the duration change in the segment. However, when speech rate changes, there are also other speech characteristics, like prosody and phoneme quality which change, and which cannot be simulated only by modifying the phoneme duration. Therefore, making speech sound natural while modifying the speech rate needs to consider and adjust these other features too.

Koreman showed \cite{Koreman_2006} that the perceived speech rate is also influenced by the listener's knowledge of the expected articulations for a particular utterance, therefore it makes a difference if all expected phones are also articulated or not. Consequently, one could compare speech rate considering only articulated vowels/syllables, as well as considering both realized and intended syllables, and test both methods with a given model. One possible result of such an approach would be the prediction of phonemes reaching $duration = 0$ at specific speech rates, which means predict phoneme drop phenomena. In spontaneous speech, the word and syllable boundaries change, and this change influences other phonetic aspects such as stress, syllable duration, phoneme duration.

@@

\vspace*{\fill}
\pagestyle{empty}
{\normalsize
\begin{center}\textbf{Eidesstattliche Erklärung}\end{center}
Hiermit versichere ich an Eides statt, dass ich die vorliegende Arbeit im Bachelorstudiengang Wirtschaftsinformatik selbstständig verfasst und keine anderen als die angegebenen Hilfsmittel – insbesondere keine im Quellenverzeichnis nicht benannten Internet-Quellen – benutzt habe. Alle Stellen, die wörtlich oder sinngemäß aus Veröffentlichungen entnommen wurden, sind als solche kenntlich gemacht. Ich versichere weiterhin, dass ich die Arbeit vorher nicht in einem anderen Prüfungsverfahren eingereicht habe und die eingereichte schriftliche Fassung der auf dem elektronischen Speichermedium entspricht.
\vspace*{1cm}\\
Hamburg, den 06.04.2017
\hspace*{\fill}\begin{tabular}{@{}l@{}}\hline
\makebox[5cm]{Vorname Nachname}
\end{tabular}
\vspace*{3cm}

>>>>

\documentclass[11pt,a4paper]{scrbook}
\usepackage{geometry}	
%\usepackage[backend=bibtex,style=numeric]{biblatex}
%\usepackage{ngerman}
%\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[style=numeric]{biblatex}
\bibliography{../PaperBib}
\usepackage{colortbl}	
\usepackage{xcolor}
%\usepackage{soul}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[outdir=../Graphen/]{epstopdf}
\usepackage{textcomp}
%\usepackage{floatrow}
\renewcommand{\figurename}{Fig.}
\usepackage[skip=0pt]{caption}
\captionsetup{format=hang}
\setcapindent{0pt}
\usepackage{subcaption}
\usepackage{scrhack}
%\renewcommand{\familydefault}{\sfdefault}
\definecolor{uhhred}{cmyk}{0,100,100,0}
\usepackage{tabularx}

@@

\section{Method}
Our starting premise was that phoneme durations are being influenced by the speaking rate. In addition to this, we assumed that this influence is not equal on all phonemes, and on all situations, and therefore not linear. So we formulated our aim as to investigate the influence of speech rate on the duration of different phonemes and find a way of using this information to improve the performance of phoneme prediction models.

In order to achieve this aim, we first examined the phonemes with respect to their duration. We found out that the classical phonetics book grouping of phonemes is not respected in terms of duration. There are consonants like \texttt{/N/}, and \texttt{/m/} which may be longer than some long vowels, and long vowels which are shorter than some short vowels or even several consonants, like \texttt{/y:/}. If one would apply a clustering method on phonemes to group them on seven classes based on their duration interval coordinates using, say k-means method, the resulting groups would certainly differ from those listed in Chapter \ref{chap_phonemes}.

Therefore we decided to capture phoneme identity in our model as well, although it inflates the number of attributes and attribute values to compute. The relation between phoneme duration and factors other than speech rate was treated only marginally, as these have already been extensively examined in the literature on various corpora. For the German language, intuition aside, we have information on such factors provided by: Kohler \cite{Kohler1992}, Riedi \cite{Riedi1995}, Moebius \cite{Moebius1996}, and Brinckmann \& Trouvain \cite{Brinckmann_2003}, who applied some of the existent segment duration models to the German language. At this stage, we considered such factors only to the extent they might have helped identify groups of phonemes that may be best suited to evaluate the influence of speech rate. Such a factor is primary stress, which creates a group of very elastic vowels, which are therefore also sensitive to changes in speech tempo.

Then we looked for a definition of speech rate that can be used to model phoneme durations. This supposed the presence of a good correlation between speech rate and phoneme duration, at least for some phoneme groups. We explained our choice whith respect to this in Chapter \ref{chap_SR}. 
The fact, that speech rate doesn't show any correlation with some phonemes, like \texttt{/r/}, while it has a pretty good correlation with others, like \texttt{/a:/} suggests the need for classification in whichever model we choose.

@@

\begin{enumerate}
	\item The first one predicted the phoneme duration as the mean of its durations across the whole corpus. 
	\item The second one built a two level tree, which split the root on the 49 realized phonemes, and then each phoneme on two specific speech rate values. The leaves contained three predicted values for each phoneme. These values were calculated using the mean duration of the specific phoneme across the corpus modified or not by a proportion of its standard deviation, based on the speech rate value split. 
\end{enumerate}

@@

Next, we pruned the used feature space by using two feature selection algorithms: forward and backward feature selection. Their principle is easy to understand and they are available in Weka. After pruning, the \textit{word} related features got completely eliminated, together with \textit{syllable-place-in-word} and \textit{articulation-of-realised-previous phoneme}. Consequently, following features were selected:

@@

	\begin{itemize}
		\item phoneme id: all 52 phoneme definitions used in Verbmobil
		\item phoneme type: vowel, or consonant
		\item articulation manner: vowel, plosive, fricative, nasal, lateral, approximant, other
	\end{itemize}

@@

	\begin{itemize}
		\item position of phoneme in syllable: initial, final
		\item syllable part containing current phoneme: onset, nucleus, coda, single-phoneme-syllable
		\item lexical stress of the syllable (using the stress type of the nucleus vowel)
	\end{itemize}

@@

	\begin{itemize}
		\item previous phoneme type: vowel, consonant, or none
		\item following phoneme type: vowel, or consonant, or none
		\item articulation manner of following phoneme: vowel, plosive, fricative, nasal, lateral, approximant, other, none
		\item if following phoneme is voiced: yes, no, or none
		\item syllable part to which following phoneme belongs: onset, nucleus, coda, single-phoneme-syllable
	\end{itemize}
\end{itemize}

Note that all features refer to actually realised units in speech. We also had to reduce the amount of data used for the model to 58\% of the total corpus, due to capacity limitations imposed by Weka on our system.

To this collection of features we added \textit{speech rate} in the second phase and compared the results.

\section{Results}
\label{chap_6:results}
We trained and evaluated both a M5P and a REPTree model. The results were very similar, so we list here only the results obtained using the model tree M5P, as they were slightly better. As can be seen in Table \ref{tab:perfM}, adding the speech rate feature caused a significant improvement on all our models.

@@

\hline
	 & Basic - SR & Basic + SR & M5P - SR & M5P + SR & CART\\
\hline
\hline
Correlation coef 		& .42  &  & .50 & .79 & .86 / .83\\
	\hline
MAE  &  31.02 & & 34.21 & 25.19 & 22.46 / 21.40\\
	\hline
RMSE  &  47.09 & & 62.91 & 44.70 & NA \\
	\hline
\end{tabular}
\vspace{4mm}
\caption{Overview of the performance results of 3 models with (+) and without (-) speech rate (SR). CART is the model created by Brinckmann \& Trouvain \cite{Brinckmann_2003}. The two values in the CART cells represent the performance obtained for 2 different speakers. RMSE was not reported for this model.} 
\label{tab:perfM}
\end{table}

However, the performance of our model tree is weaker than the performance of the CART model built by Brinckmann \& Trouvain \cite{Brinckmann_2003}. This is explainable if we consider the nature of the used datasets. The CART tree used PhonDat, a rather homogenous dataset in terms of speech rate and number of speakers. Furthermore, the model was trained and tested separately on data produced by one speaker. The reported results show even for PhonDat a difference in performance between speakers. Another aspect that can have a great impact on model performance is the quality of the input data. While PhonDat was manually segmented and annotated, Verbmobil was processed automatically using the MAUS segmentation system. When evaluating outliers we found many examples of wrong segmentation, as we also discussed in Sections \ref{Q} and \ref{cons}.

On the other hand, our model tree shows robustness with respect to factors like inter-speaker variability and speech rate. The results obtained for a test set containing data from a speaker not included in the training data were similar to those obtained on known speakers on terms of correlation coefficient, and showed less error rates, as can be seen in Table \ref{tab:perfM5P}. Moreover, if the model is to be evaluated on data produced by only one speaker, it produces good results even when using less than 10\% of the amount of data used to train on the multiple speaker corpus. The most frequent speaker in our database still only has 185 recorded turns, consisting of 9315 identifyiable uttered phonemes, which represent 0.02\% of the total phoneme instances in our corpus, so we could not test our model on a larger dataset of a single speaker. In comparison to this, the dataset used for CART contained 624 sentences, and a total of 4932 orthographically different words, which results in more than a double sized corpus.

@@

\end{tabular}
\vspace{4mm}
\caption{Overview of the performance of our M5P model tree on different training and test sets.} 
\label{tab:perfM5P}
\end{table}

\chapter{Conclusion and Outlook}
Speech tempo is an important feature of natural speech, which produces lengthening or shortening of the speech units with reference to some ``standard`` speaking tempo. Our experiment shows that we can trace this influence at phoneme level, where it produces a non-linear modification in terms of duration of individual phonemes. Moreover, we have seen that adding a feature of tempo to a phoneme duration prediction model resulted in a significant improvement of model performance, even with a significant quantity of noise in the input data, as we stated in the discussion of our experiment results in Section \ref{chap_6:results}.

The main difficulty we encountered was to find an appropriate definition of speech rate to be used in our models. Our definition considers the structure of the corpus used and the purpose it needs to serve. In addition, it did produce a significant improvement of the model used, so it may be used in phoneme duration prediction models. However, a universal definition of speech rate, applying to different corpora, would be desirable.

Future work would be to train and test such a model using the speech rate feature on data with a highly reduced amount of noise, which should result in an improved performance.

A further possible improvement may achieved considering Koreman's observation \cite{Koreman_2006} on speech rate. He showed that perceived speech rate is influenced by the listener's knowledge of the expected articulations for a particular utterance, therefore it makes a difference if all expected phones are also articulated or not. Consequently, one could compare speech rate considering only articulated vowels/syllables, as well as considering both realized and intended phonemes, testing both methods with a given model. One possible result of such an approach would be the prediction of phonemes reaching $duration = 0$ at specific speech rates, which means predicting phoneme drop phenomena. In spontaneous speech, the word and syllable boundaries change, and this change influences other phonetic aspects such as stress, syllable duration, phoneme duration.

More generally, we spoke about improving the quality of rendered speech under varying tempo. Correct phoneme length prediction is just a step when improving the speech quality retrieved by audio programs when the rendering speed is modified, which we adressed in this work. However, when tempo changes in natural speech, other speech characteristics are being modified as well, like prosody, pitch, and phoneme quality, which cannot be simulated only by modifying the phoneme duration. Therefore, making speech sound natural while modifying the rendering speed needs to consider and adjust these other features too, implying further work to investigate changes of these factors produced by speech rate modification.

@@

\vspace*{\fill}
\pagestyle{empty}
{\normalsize
\begin{center}\textbf{Eidesstattliche Erklärung}\end{center}
Hiermit versichere ich an Eides statt, dass ich die vorliegende Arbeit im Bachelorstudiengang Informatik selbstständig verfasst und keine anderen als die angegebenen Hilfsmittel – insbesondere keine im Quellenverzeichnis nicht benannten Internet-Quellen – benutzt habe. Alle Stellen, die wörtlich oder sinngemäß aus Veröffentlichungen entnommen wurden, sind als solche kenntlich gemacht. Ich versichere weiterhin, dass ich die Arbeit vorher nicht in einem anderen Prüfungsverfahren eingereicht habe und die eingereichte schriftliche Fassung der auf dem elektronischen Speichermedium entspricht.
\vspace*{1cm}\\
Hamburg, den 06.04.2017
\hspace*{\fill}\begin{tabular}{@{}l@{}}\hline
\makebox[5cm]{Vorname Nachname}
\end{tabular}
\vspace*{3cm}
